\chapter{Decidability}

\section*{Lecture 15: Encodings, Universal Turing Machine, Countable Sets}

\emph{Encodings} \textsf{TM} input is always a string. Need to represent other objects as strings. Same as ``serialization'', e.g. JSON. If we have an object $O$ (e.g. a graph) then $\encoding{O}$ represents it as a string. Assume TMs can do serialization.

We can also represent TMs as strings.

\begin{equation*}
    HALT_{\text{TM}} = \set{\encoding{M, w}: M \text{ is a TM, } w \in \Sigma^* \text{, and } M \text{ halts on input} w} \tag{The Halting Problem}
\end{equation*}

\begin{equation*}
    A_{\text{TM}} = \set{\encoding{M, w}: M \text{ accepts } w}
\end{equation*}

What strings are in $A_{\text{TM}}$!

\begin{enumerate}
    \item If $x$ is of form $\encoding{M, w}$ and $M$ accepts $w$, then $x \in A_{TM}$ (or rejects for $HALT_{TM}$)
    \item If $x$ is of form $\encoding{M, w}$ and $M$ rejects for $HALT_{TM}$ or loops on input $w$, then $x \notin A_{TM}$
    \item If $x$ \emph{not} of form $\encoding{M, w}$ then $x \notin A_{TM}$ (Junk)
\end{enumerate}

Let's show that $A_{TM}$ is \emph{recognizable}. We need to design a TM s.t. on input $x$, if $x$ is in:

Case 1: our TM should accept $x$

Case 2/3: TM can reject/run forever

Idea(Turing): A TM can simulate another TM given via an encoding in input.

$U$ is a TM that recognizes $A_{TM}$, but it is not a decider. Conclusion: $A_{TM}$ is recognizable.

\begin{definition}
    Two sets $A$ \& $B$ have the same cardinality if there exists a bijection $f: A \rightarrow B$.
\end{definition}

\begin{definition}
    A set is called \emph{finite} if it has a bijection to $\set{1, \ldots, n}$ for some natural number $n$.
\end{definition}

\begin{definition}
    A set is \emph{countably infinite} if it has the same cardinality as $\N$.
\end{definition}

\begin{definition}
    A set is \emph{countable} (or \emph{at most countable}) if it is finite or countably finite.
\end{definition}

Let $E: \set{n: n \in \N \text{ and } n \text{ is even}}$

\emph{Claim}: $E$ is countably infinite.

\begin{proof}
    Need a bijection $f: \N \rightarrow E$. Can use $f(x) = 2x$.
\end{proof}

\section*{Lecture 16: Undecidability of ATM}

$A$ =  statement on the left is False. $B$ = statement on the right is True. If $A$ is true $\Rightarrow$ B is false $\Rightarrow$ $A$ is false.

$\set{0, 1}^* = \set{\epsilon, 0, 1, 00, 01, \ldots}$ countable

$B = \set{0000\ldots, 111\ldots, 010101\ldots}$

Why not apply diagonalization to $\set{0, 1}^*$?

We need to define $a_n$. Each row $f(n)$ has finite length. If $f(n)$ has finite length $\geq n$, then $a_n$ is undefined.

$2^{\Sigma^*}$ is uncountably infinite.

There are many TMs

\begin{description}
    \item $\Rightarrow$ countably many decidable many languages (each decided by some TM)
    \item $\Rightarrow$ countably many recognizable languages (each recognized by some TM)
\end{description}
