\chapter{Reducibility}

\section*{Lecture 17: Reductions}

Recall:

\begin{itemize}
    \item $A_{\text{TM}} = \set{\encoding{M, w} : M \text{ accepts } w}$
    \item $HALT_{\text{TM}} = \set{M \text{ halts on } w}$
\end{itemize}

\begin{theorem}
    $HALT_{\text{TM}}$ is undecidable.
\end{theorem}


\begin{proof}
    Last time showed that $A_{\text{TM}}$ is undecidable. Show that $A_{\text{TM}} \leq_{T} HALT_{\text{TM}}$. Suppose we have a TM $R$ that decides $HALT_{\text{TM}}$. Then we want to create a TM $S$ that decides $A_{\text{TM}}$.

  Design $S$: On input $x$,

  \begin{enumerate}
        \item Reject if $x$ not in form of $\encoding{M, w}$.
        \item Run $R$ on input $\encoding{M, w}$.
        \item If $R$ accepts, (we know $M$ halts on input $w$), simulate $M$ on input $w$. Accept if $M$ accepts, reject if $M$ rejects. Else ($M$ runs forever on input $w$). Reject. ($M$ does not accept $w$)
  \end{enumerate}

  So $S$ is a decider for $A_{\text{TM}}$. Contradiction!
\end{proof}

\emph{Claim}: Suppose $L$ and $\overline{L}$ are both recognizable, then $L$ is decidable (and so is $\overline{L}$.

\begin{proof}
    Let $M_1$ be TM that recognizes $L$. Let $M_2$ be TM that recognizes $\overline{L}$. Design a new TM $M_3$ as follows: on input $x$,

  \begin{enumerate}
        \item In parallel simulate both $M_1$ and $M_2$.
        \item If $M_1$ halts and accepts then $M_3$ accepts.
        \item If $M_2$ halts and accepts then $M_3$ rejects.
  \end{enumerate}

    Suppose $x \in L$. Then $M_1$ eventually accepts $x \Rightarrow M_3$ accepts $x$. $x \notin L$, then $M_2$ eventually accepts $x \Rightarrow M_3$ rejects $x$. So $M_3$ decides $L$.
\end{proof}

\begin{corollary}
    $\overline{A_{\text{TM}}}$ is not recognizable.
\end{corollary}

\begin{proof}
    If it was recognizable, Claim should imply $A_{\text{TM}}$ is decidable!
\end{proof}

\section*{Lecture 18: More reductions}

\begin{equation*}
    E_{TM} = \{\langle N \rangle : N \text{ is a TM s.t. } L(N) = \emptyset\}
\end{equation*}

\begin{theorem}[Sipser 5.1]
    $E_{TM}$ is undecidable.
\end{theorem}

\begin{proof}
    We know $A_{\text{TM}}$ is undecidable. We want to prove $A_{\text{TM}} \leq_{T} E_{TM}$. Then we can conclude that $E_{\text{TM}}$ is undecidable.
    How to do reduction? Assume that $R$ is a (hypothetical) TM solving $E_{\text{TM}}$. We must somehow construct a TM $S$ that decides $A_{TM}$. $A_{TM} = \{\encoding{M, w}: M \text{ accepts } w\}$.

    \emph{Design of $S$}:

    On input $X$:
    \begin{enumerate}
        \item If $x$ is not of form $\encoding{M, w}$, reject.
        \item Construct description of $N$ like this:

        Let $N$ be TM: On input $y$, (ignore input), simulate $M$ on input $w$:

        \begin{enumerate}
            \item If $M$ accepts $N$ accepts.
            \item If $M$ rejects $N$ rejects.
        \end{enumerate}

        \item Run $R$ on input $\encoding{N}$.
        \item Accept if $R$ rejects; reject if $R$ accepts.
    \end{enumerate}

    $S$ is a decider since $R$ is.

    \emph{What is $L(N)$}:

    Case 1: $M$ accepts $w$. Then $L(N) = \Sigma^*$.

    Case 2: $M$ does not accept $w$:

    \begin{enumerate}[2a.]
        \item $M$ rejects $w$. Then $N$ rejects all inputs, so $L(N) = \emptyset$.
        \item $M$ runs forever on input $w$. Then $N$ runs forever on all inputs, so $L(N) = \emptyset$.
    \end{enumerate}

    \emph{Analysis}:

    In Case 1, then $R$ rejects $\encoding{N}$, so $S$ accepts ($M$ accepts $w$). Otherwise, $L(N) = \emptyset$, so $R$ accepts $\encoding{N}$, so $S$ rejects.
\end{proof}

\begin{equation*}
    REGULAR_{\text{TM}} = \set{\encoding{N} : N \text{ is a TM s.t. } L(N) \text{ is regular}}
\end{equation*}

\begin{proof}
    We know $A_{\text{TM}}$ is undecidable. We want to prove $A_{\text{TM}} \leq_{T} E_{\text{TM}}$. Then we can conclude that $REGULAR_{\text{TM}}$ is undecidable.

    \emph{What is $L(N)$}:

    Case 1: $M$ accepts $w$. Then $L(N) = \set{0^n1^n : n \in \N}$ (NOT REGULAR!).

    Case 2: $M$ does not accept $w$:

    \begin{enumerate}[2a.]
        \item $M$ rejects $w$. $L(N) = \emptyset$ is regular.
        \item $M$ runs forever on input $w$. $L(N) = \emptyset$ is regular.
    \end{enumerate}

    \emph{Analysis}:

    If $M$ accepts $w$, then $L(N)$ is not regular, so $R$ rejects, so $S$ accepts. Otherwise, $L(N) = \emptyset$, so $R$ accepts $\encoding{N}$, so $S$ rejects.
\end{proof}

